---
title: "Lecture 123 Supplement"
author: "Katy McKeough"
date: "1/31/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gam in Action

```{r cars}
library(gam)
data(kyphosis)
?kyphosis
```

General Additive Models allow us to model binary variables. Let us model the response variable *Kyphosis* which indicates whether or not the condition is present or not. Use *Age*, *Number* and *Start* as predictor variables.

```{r pressure}
kyph1.gam = gam(Kyphosis ~ s(Age) + s(Number) + s(Start),
  family=binomial, data=kyphosis)
print(summary(kyph1.gam))

par(mfrow=c(1,3))
plot(kyph1.gam, resid=T, se=T, rug=F, pch=19, col="red")
```

Perform a likelihood ratio test against the model with no terms. Is incorporating the three variables better than nothing?

```{r}
kyph0.gam = gam(Kyphosis ~ 1,
  family=binomial, data=kyphosis)

print(anova(kyph0.gam,kyph1.gam,test='Chi'))

```


It appears individual effects are not significant in presence of others. Remove Number and re fit the model

```{r}
kyph2.gam = gam(Kyphosis ~ s(Age) + s(Start),
  family=binomial, data=kyphosis)
print(summary(kyph2.gam))

par(mfrow=c(1,2))
plot(kyph2.gam, rug=T, se=T)

```

Visualy, t is clear that the response does not behave linearly in terms of these predictors. Here we fit the corresponding GLM and test the comparison to the GAM.

```{r}
kyph2.glm = glm(Kyphosis ~ Age + Start,
  family=binomial, data=kyphosis)
print(anova(kyph2.glm,kyph2.gam,test="Chi"))
```

We can use GAMs for prediction the same as we have done before:

```{r}
kyph.new = data.frame(Age = c(84,85,86), Start=c(7,8,9))
print(predict(kyph2.gam,kyph.new,type="response"))
```


Another use of GAM is to transform predictors for use in linear models. Based on the plots for *Age* and *Start* it makes sense to include quadratic terms.
```{r}
kyphosis$Age.sq = (kyphosis$Age - mean(kyphosis$Age))^2
kyphosis$Start.sq = (kyphosis$Start - mean(kyphosis$Start))^2
kyph3.glm = glm(Kyphosis ~ Age + Start + Age.sq + Start.sq,
  family=binomial, data=kyphosis)

print(summary(kyph3.glm))

print(anova(kyph2.glm,kyph3.glm,test="Chi"))
```



