---
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

# load libraries and set bayesplot theme 
# make sure you have these packages installed
```{r load-packages, message=FALSE, warning=FALSE}
library("rstan")
library("ggplot2")
library("bayesplot")

theme_set(bayesplot::theme_default())
```


# Load and examine data
```{r count-data, fig.width = 4, fig.height = 4, message=FALSE}
# Loads vector of counts 'y'
source("count-data.R")

N <- length(y)
print(N)
print(y)
qplot(y)
```

# Compare our data to draws from Poisson with same mean
```{r plot-x, fig.width = 4, fig.height = 4, message=FALSE}
x <- rpois(N, lambda = mean(y))
qplot(x)
```

```{r plot-y-x, message=FALSE}
plotdata <- data.frame(
  value = c(y, x), 
  variable = rep(c("Our data", "Poisson data"), each = N)
)

# Frequency polygons
ggplot(plotdata, aes(x = value, color = variable)) + 
  geom_freqpoly(binwidth = 0.5) +
  scale_x_continuous(name = "", breaks = 0:max(x,y)) +
  scale_color_manual(name = "", values = c("gray30", "purple"))

# Side-by-side bar plots
ggplot(plotdata, aes(x = value, fill = variable)) + 
  geom_bar(position = "dodge") +
  scale_x_continuous(name = "", breaks = 0:max(x,y)) +
  scale_fill_manual(name = "", values = c("gray30", "purple"))
```

# Fit basic Poisson model

Even though we already suspect it won't be a good model for this data, it's
still a good idea to start by fitting the simplest Poisson model. From there we
can then identify in which ways the model is inadequate.
```{r create_list, fig.width = 4, fig.height = 4, message=FALSE}
# create list 
stan_list <- list()
stan_list$Y <- y
stan_list$N <- N
```

```{r stan_code, fig.width = 4, fig.height = 4, message=FALSE}
# stan code 
stan_code <- c("
data {
  int N;                // Number of observations
  int<lower=0> Y[N];    // Count data (integer array)
}

parameters {
  real<lower=0> lambda; // Poisson rate parameter 
}

model {
  // Prior
  lambda ~ exponential(0.1);

  // Likelihood
  Y ~ poisson(lambda);
}

generated quantities {
  int y_rep[N];         // Draws from posterior predictive dist
  for (n in 1:N) {
    y_rep[n] = poisson_rng(lambda);
  }
}

")
```

```{r fit_model, fig.width = 4, fig.height = 4, message=FALSE}
# fit the model 
options(mc.cores = parallel::detectCores())
fit <- stan(model_code = stan_code, 
            data = stan_list, 
            iter = 2000, 
            chains = 4)
```

```{r, plot_convergence}
plot(fit, plotfun="trace", pars='lambda')
```

# Look at posterior distribution of lambda
```{r, plot_lambda}
color_scheme_set("brightblue") # check out bayesplot::color_scheme_set
lambda_draws <- as.matrix(fit, pars = "lambda")
mcmc_areas(lambda_draws, prob = 0.95) # color 95% credible interval
```

# Compare posterior of lambda to the mean of the data
```{r, print_fit}
means <- c("Posterior mean" = mean(lambda_draws), "Data mean" = mean(y))
print(means, digits = 3)
```
The model gets the mean right, but, as we'll see next, the model is quite bad
at predicting the outcome.

# Graphical posterior predictive checks
# Extract `y_rep` draws from the fitted model object
```{r y_rep}
y_rep <- as.matrix(fit, pars = "y_rep")
```

# Compare histogram of `y` to histograms of several `y_rep`s
```{r ppc-hist, message=FALSE}
ppc_hist(y, y_rep[1:8, ], binwidth = 1)
```

# Compare density estimate of `y` to density estimates of a bunch of `y_rep`s
```{r ppc-dens-overlay}
ppc_dens_overlay(y, y_rep[1:50, ])
```

# clear environment and load normal data
```{r load_normal_data}
rm(list = ls())
mydata <- read.csv("practice_data.csv")
```


```{r examine_normal_data}
X <- mydata$x
N <- length(X)
print(mean(X))
qplot(X)
```

```{r fake_normal_data, fig.width = 4, fig.height = 4, message=FALSE}
Y <- rnorm(N, mean=mean(X), sd=sd(X))
qplot(Y)
```

```{r normal_side_by_side, fig.width = 4, fig.height = 4, message=FALSE}
plotdata <- data.frame(
  value = c(X, Y), 
  variable = rep(c("Our data", "Normal data"), each = N)
)

ggplot(plotdata, aes(x=value)) + 
    geom_histogram(data=subset(plotdata, variable == 'Our data'),
                   fill = "red", alpha = 0.25) +
    geom_histogram(data=subset(plotdata, variable == 'Normal data'),
                   fill = "blue", alpha = 0.25) 
```

```{r create_normal_list, fig.width = 4, fig.height = 4, message=FALSE}
# create list 
stan_list <- list()
stan_list$X <- X
stan_list$N <- N
```

```{r normal_stan_code, fig.width = 4, fig.height = 4, message=FALSE}
# stan code 
stan_code <- c("
data {
  // write code here
}

parameters {
 // write code here
}

model {
 // write code here
}

generated quantities {
 // (optional) write code here
}

")
```

```{r fit_normal_model, fig.width = 4, fig.height = 4, message=FALSE}
# fit the model 
options(mc.cores = parallel::detectCores())
fit <- stan(model_code = stan_code, 
            data = stan_list, 
            iter = 2000, 
            chains = 4)
```




